---
title: "Customer Churn"
author: "Anthony Bouz"
date: "12-06-2023"
output:
  word_document: default
  pdf_document: default
---

### Introduction
Customer churn, also known as customer attrition, occurs when customers stop doing business with a company or stop using a company's services. By being aware of and monitoring churn rate, companies are equipped to determine their customer retention success rates and identify strategies for improvement. We will use a machine learning model to understand the precise customer behaviors and attributes which signal the risk and timing of customer churn. In this project, I want to predict Telco customer churn based on their behavior. This dataset is taken from Kaggle website. The dataset being used can be accessed at "https://www.kaggle.com/datasets/blastchar/telco-customer-churn/data". I will be using logistic regression and Gradient Boosting to predict the Churn Value variable that helps in analysing accuracy,specificity,sensitivity.

### Load the data
```{r echo=TRUE , warning= FALSE , message= FALSE}
library(tidyverse)
library(gridExtra)
library(readxl)
churn <- read_xlsx("Telco_customer_churn.xlsx")
```

### Exploratory data analysis
```{r}
str(churn)
```

The churn dataset has 7043 observations and 33 variables

### check missing values
```{r}
sapply(churn, function(x) sum(is.na(x)))
```
There are 5174 missing values in Churn Reason variable which is more than 50% so I will be discarding the variable and since there are only 11 missing in Total Charges, I will be removing those missing values


##Remove missing values
```{r}
churn = churn %>% select(-`Churn Reason`)
churn <- na.omit(churn)
```

```{r}
churn<- churn %>% select(-c(CustomerID,Count,Country,State,City,`Zip Code`,`Lat Long`,Latitude,Longitude,`Churn Label`))
```

CHurn Label is removed as we have a variable called Churn Value which is in binary and represents the same, the rest are being removed as they don't add any value while building the model on Churn.

In churn value 1 represents yes the customer churned and 0 represents No

```{r}
churn <- churn %>% mutate_if(is.character , as.factor)
str(churn)
```

```{r}
churn$`Churn Value` = as.factor(churn$`Churn Value`)
ggplot(churn, aes(x=`Churn Value`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count..-200, label = paste0(round(prop.table(..count..),4) * 100, '%')), stat = 'count', 
position = position_dodge(.1), size = 3)+ ylab("Count")
```

From the graph we see there are 26.58% customers that churn

```{r}
churn %>% 
  group_by(`Tenure Months`, `Churn Value`) %>% 
 dplyr::summarise(Number = n()) %>% 
  ggplot(aes(`Tenure Months`, Number)) +
  geom_line(aes(col = `Churn Value`)) +
  labs(x = "Tenure (month)",
       y = "Number of Customer",
       title = "Churn Based on Tenure") +
  scale_x_continuous(breaks = seq(0, 70, 10)) +
  theme_minimal()
```

From the graph we see that Customers are likely to churn after 1 month.


```{r}
churn_score = churn %>% 
  group_by(`Churn Value`) %>% 
  summarise(Churn_Score = mean(`Churn Score`))

ggplot(churn_score, aes(x = `Churn Value`, y = Churn_Score, fill = `Churn Value`)) +
  geom_bar(stat = "identity") +  # Use geom_bar for bar plot
  geom_text(aes(label = round(Churn_Score, 2)),  # Use round to limit decimal places
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Average Churn Score by Churn Value",
       x = "Churn Value",
       y = "Average Churn Score") +
  theme_minimal()
```
Churn Score of 82.51043 has a high likelihood of churning (Churn Value = 1), while a customer with a Churn Score of 50.10149 is less likely to churn (Churn Value = 0).

```{r}
cltv_score = churn %>% 
  group_by(`Churn Value`) %>% 
  summarise(CLTV_Score = mean(`CLTV`))

ggplot(cltv_score, aes(x = `Churn Value`, y = CLTV_Score, fill = `Churn Value`)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(CLTV_Score, 2)),
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Average CLTV by Churn Value",
       x = "Churn Value",
       y = "Average CLTV") +
  theme_minimal()
```
On average, customers who remained with the company (Churn Value = 0) have a higher CLTV compared to those who left (Churn Value = 1).


#Gender , Senior Citizen , Partner , Dependents
```{r}
a<-ggplot(churn, aes(x=Gender,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200, label = paste0(round(prop.table(..count..),4) * 100, '%')), stat = 'count', 
 position = position_dodge(.1),  size = 3)+ ylab("Count")


b<-ggplot(churn, aes(x=`Senior Citizen`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200, label = paste0(round(prop.table(..count..),4) * 100, '%')), 
 stat = 'count', position = position_dodge(0.1), size = 3)+ ylab("Count")



c<-ggplot(churn, aes(x=Partner,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200, label = paste0(round(prop.table(..count..),4) * 100, '%')), stat = 'count', 
position = position_dodge(.1), size = 3)+ ylab("Count")


d<-ggplot(churn, aes(x=Dependents,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')), 
stat = 'count',position = position_dodge(.1), size = 3) + ylab("Count")

grid.arrange(a,b,c,d,ncol=2)

```

From the gender, male and female seems to have same proportion to churn.
People who are not senior citizens are most likely to churn
Customer without partner are more likely to churn
Customer without Dependents are more likely to churn


```{r}
grid.arrange(ggplot(churn, aes(x=`Internet Service`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')), stat = 'count', position = position_dodge(.1),size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Online Security`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')),stat = 'count',position = position_dodge(0.2),size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Online Backup`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')),stat = 'count',position = position_dodge(0.2),size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Device Protection`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')),stat = 'count',position = position_dodge(0.2),size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Phone Service`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')),stat = 'count',position = position_dodge(0.2),size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Multiple Lines`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')), stat = 'count',position = position_dodge(0.2), size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Tech Support`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')), stat = 'count', position = position_dodge(0.2),size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Streaming TV`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')),stat = 'count',position = position_dodge(0.2), size = 3)+ ylab("Count"),
             
             ggplot(churn, aes(x=`Streaming Movies`,fill=`Churn Value`))+ geom_bar() + geom_text(aes(y = ..count.. -200,label = paste0(round(prop.table(..count..),4) * 100, '%')), stat = 'count',position = position_dodge(0.2),size = 3)+ ylab("Count"))
```

Customer that uses fiber optic in internet service are more likely to churn
Customer that doesn't use online security, online backup, device protection, streaming TV, and tech support are more likely to churn 

### Statistical tests
```{r}
str(churn)
chisq.test(churn$`Churn Value`, churn$Gender)
chisq.test(churn$`Churn Value`, churn$`Senior Citizen`)
chisq.test(churn$`Churn Value`, churn$Partner)
chisq.test(churn$`Churn Value`, churn$Dependents)
chisq.test(churn$`Churn Value`, churn$`Phone Service`)
chisq.test(churn$`Churn Value`, churn$`Multiple Lines`)
chisq.test(churn$`Churn Value`, churn$`Internet Service`)
chisq.test(churn$`Churn Value`, churn$`Online Security`)
chisq.test(churn$`Churn Value`, churn$`Online Backup`)
chisq.test(churn$`Churn Value`, churn$`Device Protection`)
chisq.test(churn$`Churn Value`, churn$`Tech Support`)
chisq.test(churn$`Churn Value`, churn$`Streaming TV`)
chisq.test(churn$`Churn Value`, churn$`Streaming Movies`)
chisq.test(churn$`Churn Value`, churn$`Paperless Billing`)
chisq.test(churn$`Churn Value`, churn$`Payment Method` )
chisq.test(churn$`Churn Value`, churn$Contract)

t.test(`Tenure Months`~`Churn Value`,data=churn)
t.test(`Total Charges`~`Churn Value`,data=churn)
t.test(`Monthly Charges`~`Churn Value`,data=churn)
t.test(`Churn Score`~`Churn Value`,data=churn)
t.test(CLTV~`Churn Value`,data=churn)
```

We can see that Gender and Phone Service has p value greater than 0.05 and null hypothesis can not be rejected and hence we will be discarding these 2 columns

```{r}
library(GGally)
ggcorr(churn, label = T)
```

Since Total charges has strong correlation with tenure months, Monthly Charges,Churn Score and CLTV, it can be taken out

```{r}
churn <- churn %>% select(-c(`Total Charges`,Gender,`Phone Service`))
churn$`Churn Value` = as.factor(ifelse(churn$`Churn Value` == 0 , "No","Yes"))
library(caret)
split_train_test <- createDataPartition(churn$`Churn Value`,p=0.7,list=FALSE)
set.seed(1234)
dtrain<- churn[split_train_test,]
dtest<- churn[-split_train_test,]


mod <- glm(`Churn Value` ~., data = dtrain,
              family=binomial(link='logit'))
summary(mod)
```

Here we can see that 7 coeffiencients are not defined and throws NA because the model is recognizing that two or more of the predictors are identical, or perfectly predicted by the combination of the other two predictors. This means that we cannot include all three terms in your model.


```{r}
churn <- churn %>%
  mutate(
    `Multiple Lines` = as.factor(case_when(
      `Multiple Lines` == "No phone service" ~ "No",
      TRUE ~ as.character(`Multiple Lines`)
    )),
    `Online Security` = as.factor(case_when(
      `Online Security` == "No internet service" ~ "No",
      TRUE ~ as.character(`Online Security`)
    )),
    `Online Backup` = as.factor(case_when(
      `Online Backup` == "No internet service" ~ "No",
      TRUE ~ as.character(`Online Backup`)
    )),
    `Device Protection` = as.factor(case_when(
      `Device Protection` == "No internet service" ~ "No",
      TRUE ~ as.character(`Device Protection`)
    )),
    `Tech Support` = as.factor(case_when(
      `Tech Support` == "No internet service" ~ "No",
      TRUE ~ as.character(`Tech Support`)
    )),
    `Streaming TV` = as.factor(case_when(
      `Streaming TV` == "No internet service" ~ "No",
      TRUE ~ as.character(`Streaming TV`)
    )),
    `Streaming Movies` = as.factor(case_when(
      `Streaming Movies` == "No internet service" ~ "No",
      TRUE ~ as.character(`Streaming Movies`)
    ))
  )


```


```{r}
split_train_test <- createDataPartition(churn$`Churn Value`,p=0.7,list=FALSE)
set.seed(1234)
dtrain<- churn[split_train_test,]
dtest<- churn[-split_train_test,]


mod_log <- glm(`Churn Value` ~., data = dtrain,
           family=binomial(link='logit'))
summary(mod_log)
```

#we see that Tenure length, contract status, paperless Billing and  total charges have the lowest p-values and can be identified as the best predictors of customer churn

#Confusion matrix
```{r}
lr_prob1 <- predict(mod_log, dtest,type = "response")
lr_pred1 <- ifelse(lr_prob1 > 0.6,"Yes","No")
cm_log = confusionMatrix(as.factor(lr_pred1), reference = dtest$`Churn Value`)
```

#Gradient Boosted Tree
```{r}
set.seed(1)

gbm <- train(`Churn Value`~., data = dtrain, method = "gbm",metric = "ROC", trControl = trainControl("cv", number = 10,sampling = "smote",summaryFunction = twoClassSummary,classProbs = TRUE),  preProc="nzv")
predictions <- predict(gbm,dtest)
cm_gbm = confusionMatrix(predictions,dtest$`Churn Value`)
```
### Results
```{r}
data.frame(
  Model = c("Logistic Regression", "GBM"),
  Accuracy = c(0.9137, 0.935),
  Sensitivity = c(0.9664, 0.9444),
  Specificity = c(0.7679, 0.9089),
  Kappa = c(0.7684, 0.8367)
)
```

##Conclusion

We can see from the above results Gradient Boosting model is well balanced and it's ability to correctly identify negative instances makes it an obvious choice to consider.
